---
title: "gene growth"
author: "joyfaith onoh"
date: '2022-06-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```








```{r, }
URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/gene_expression.tsv"
download.file(URL,destfile = "gene expression")
```
```{r}
gene <- read.table(URL,stringsAsFactors = FALSE, header=TRUE, row.names = 1)
head (gene)
```
```{r,}
grep("gene identifiers", rownames(gene))
head(gene)
```
```{r}
#Q2
##MAKE A NEW COLUMN WHICH IS THE MEAN OF THE OTHER COLUMNS. SHOW A TABLE OF VALUES FOR THE FIRST SIX GENES.


rowMeans(gene)# mean of the genes
Mean <- rowMeans(gene)# assining it a header as mean
Mean
gene$Mean <- Mean

head(gene)
```
#Q3 LIST THE 10 GENES WITH THE HIGHEST MEAN EXPRESSION
```{r }
order(gene$Mean) #first i order the gene
gene[order(-gene$Mean),]
```

```{r, }
Jsort <- gene[order(-gene$Mean),] #then sort it (the negative sign is to order it in descending order from the highest to 
Jsort[1:10,]
```
#Q4
## DETERMINE THE NUMBER OF GENES WITH A MEAN <10
```{r}
#DETERMINE THE NUMBER OF GENES WITH A MEAN <10
length (gene$Mean < 10)
```
 #5 MAKE A HISTOGRAM PLOT OF THE MEAN VALUES

```{r,}
hist(gene$Mean, xlab="MEAN OF GENES(cm)", ylab= "FREQUENCY", main= "HISTOGRAM OF MEAN THE DATASET GENE_EXPRESSION")
```
#Q6
#IMPORT THE CSV FILE INTO R

```{r,}
URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/growth_data.csv"
download.file(URL,destfile = "growth_data")
data<- read.csv(URL)
```
```{r}
head(data)
```
```{r}
str (data)
```

#TO GET THE COLUMN NAMES

```{r}
colnames(data)
```
# The column names are= "Site"            "TreeID"          "Circumf_2005_cm" "Circumf_2010_cm" "Circumf_2015_cm" "Circumf_2020_cm"

#Q2
#CALCULATE THE MEAN AND STANDARD DEVIATION OF TREE CIRCUMFERENCE AT THE START AND END OF THE STUDY AT BOTH SITES

```{r}
mean_start <- mean(data$Circumf_2005_cm)
mean_start

mean_end <- mean(data$Circumf_2020_cm)
mean_end
```

```{r }
sd_start <- sd(data$Circumf_2005_cm)
sd_start

sd_end <- (data$Circumf_2020_cm)
sd_end
```
#Q8
#MAKE A BOXPLOT OF TREE  CIRCUMFERENCE AT THE START AND END OF THE STUDY

```{r}
boxplot(data$Circumf_2005_cm, data$Circumf_2020_cm, main= "BOXPLOT OF TREE CIRCUMFERENCE IN 2005 AND 2020", ylab= "TREE CIRCUMFERENCE (cm)")
```

#Q9
#CALCULATE THE MEAN GROWTH OVER THE LAST 10 YEARS AT EACH SITE.
#the mean in the past 10years is from 2010 t0 2020
#I will find the mean for each for 2010 and 2020 and subtract

```{r }
mean(data$Circumf_2010_cm)
```
```{r}
mean(data$Circumf_2020_cm)
```
```{r}
49.912 - 10.697
```
#The mean growth is =39.215cm.

